---
title: 'STAT6306: Homework 0'
subtitle: Introduction to R and Document Preparation
output:
  html_document: default
  pdf_document: default
---

# Problem 0
R is a standard software interface for computing and graphics and Rstudio is an integrated
development environment (IDE) for R.  Install both on your computer.

* R: http://lib.stat.cmu.edu/R/CRAN/
* Rstudio: https://www.rstudio.com/products/rstudio/#Desktop

Also, see the R ``tip o' the day'' document: TOTD.pdf

# Problem 1

Suppose we have the following matrix:

```{r}
set.seed(1)
A = matrix(rnorm(4*3),nrow=4,ncol=3)
```
We want to get the column mean for each column of the matrix $A$.  Do this using each of the following techniques:

### Part a
Hard coding  (that is, write $(A[1,1]+ A[2,1] + ...)/4, ...$  )

```{r}
#SOLUTION
```

### Part b
For loop(s)
```{r}
#SOLUTION
```

### Part c
The apply (or related) function 
```{r}
#SOLUTION
```

# Problem 2

Many statistical methods can be computed/analyzed using the Singular Value
Decomposition (SVD)

```{r}
set.seed(10)
X = matrix(rnorm(24),nrow=6,ncol=4)
X[,1] = 1
```

### Part a.  

What role does X[,1] = 1 play in the above code?

#### Solution

### Part b
We can always write the unique docomposition: $X = UDV^{\top}$.  We
can compute it using R:

```{r}
svd.out = svd(X)
```

Demonstrate that matrices $U$ and $V$ are orthogonal via matrix multiplication.

#### Solution