---
title: "R and SQL"
author: "Darren Homrighausen"
date: "10/06/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction
These notes give an introduction to using SQL from within R.
Data that looks like spreadsheets/matrices are known as a **flat databases** or **flat files**.
SQL is a standard interface for storing/accessing **relational databases**.


# Relational databases

Relational databases are a common way for information to be stored and distributed.
In some cases, relational structures allow for much less redundancy.  For instance,
suppose I have a database with information about two different sections of a statistics class, each
containing 3 students, and their (Post test) - (Pre test) scores.

Using a relational database, I would have two tables:
```{r,echo=FALSE,results='asis'}
require(knitr)
tab1 = data.frame('ID'=c('s_101_1a','s_101_1b','s_101_1c'),
                 'diff_score' = c(12,5,9))
tab2 = data.frame('ID'=c('s_101_2a','s_101_2b','s_101_2c'),
                 'diff_score' = c(4,-3,6))
kable(tab1,caption='STAT101_1')
kable(tab2,caption='STAT101_2')
```

Converting this to a flat file would entail:

```{r,echo=FALSE,results='asis'}
tab3 = data.frame('ID'=c('s_101_1a','s_101_1b','s_101_1c','s_101_2a','s_101_2b','s_101_2c'),
                  'Class' = c(rep('STAT101_1',3),rep('STAT101_2',3)),
                 'diff_score' = c(12,5,9,4,-3,6))
kable(tab3)
```

Notice that even in this tiny example, there is 1.5 times as much information stored, much of
it redundant.

## SQL in R
When *RSQLite* is installed, *SQLite* comes with it.  There is another package *sqldf* 
which loads/downloads the relevant parts of both *RSQLite* and *SQLite*.  This is what
we are going to be referencing.  Install it first (install.packages('sqldf')) and then load it into memory:


```{r}
  require('sqldf')
#Note: if you get an error like:
#  unable to load shared object '/Library/Frameworks/R.framework/Resources/modules//R_X11.so':
#Then, at least on Mac OS, you need to reinstall "XQuartz", which builds X11: https://www.xquartz.org/
#You'll need to log out/in and then open/close XQuartz (this re-builds the symlinks that the warning
#                                                       is referring to)
```

 RSQLite’s *dbConnect()* function opens a connection to a database. 
 If the named database does not yet exist, one is created. The command 
 below opens a connection to the *example.sqlite* database. 
 If the database does not yet exist, one is created in R’s working directory
 
```{r}
db = dbConnect(SQLite(), dbname='example.sqlite')
getwd()
'example.sqlite' %in% dir()#This checks if example.sqlite is in cwd
```

Technically, a database hasn't been created, yet. Merely a connection
to a place on the harddrive has been opened, along with a label for
that connection.  First, we
have to make sure there isn't a Table with the same name 
(for instance, if we have run the code multiple times)
```{r}
if('STAT101_1' %in% dbListTables(db)){dbRemoveTable(db, 'STAT101_1')}
if('STAT101_2' %in% dbListTables(db)){dbRemoveTable(db, 'STAT101_2')}
```
Now, let's put in the column names and data types via a **SQL query** :
```{r}
dbSendQuery(conn = db,
       'CREATE TABLE STAT101_1
       (ID TEXT,Diff_score INTEGER)')
```
Now, we have a SQL database initialized, but it still doesn't have
any data.  First, we will manually add some observations

```{r}
dbSendQuery(conn = db,
         "INSERT INTO STAT101_1
         VALUES ('s_101_1a',12)")
dbSendQuery(conn = db,
         "INSERT INTO STAT101_1
         VALUES ('s_101_1b',5)")
dbSendQuery(conn = db,
         "INSERT INTO STAT101_1
         VALUES ('s_101_1c',9)")
```
Let's inspect this database to see what it looks like.
Note that you wouldn't actually print out the data
from a database in practice as it would be too much output.
```{r}
dbListTables(db)
dbListFields(db, 'STAT101_1')
dbReadTable(db, 'STAT101_1')
dbDisconnect(db)
```

Likewise, we could create a second SQL table that has the information
from STAT101_2.

In general, this procedure for populating the database is
too cumbersome/error prone to use in practice.  Instead, we can
directly load existing (separate) flat files in databases.
First, let's pretend we have two files on our hard drives
from the two class (practically speaking, we are creating those
files now)


```{r}
tab1 = data.frame('ID'=c('s_101_1a','s_101_1b','s_101_1c'),
                 'diff_score' = c(12,5,9))
write.csv(tab1,file='STAT101_1.csv',row.names = FALSE)
tab2 = data.frame('ID'=c('s_101_2a','s_101_2b','s_101_2c'),
                 'diff_score' = c(4,-3,6))
write.csv(tab2,file='STAT101_2.csv',row.names = FALSE)
```

Let's load in the data from csv files.  Again, we need to make sure
the table names are unused (say, if this code has been run multiple 
times.  Alternatively, you can use the 'overwrite = TRUE' 
parameter in *dbWriteTable* to, you guessed it, overwrite any 
existing information.  Likewise, if you have a large flat file,
you can load it in chunks and add it to the database via
'append = TRUE'):
```{r}
db = dbConnect(SQLite(), dbname='example.sqlite')
if('STAT101_1' %in% dbListTables(db)){dbRemoveTable(db, 'STAT101_1')}
if('STAT101_2' %in% dbListTables(db)){dbRemoveTable(db, 'STAT101_2')}
```

```{r}

STAT101_1 = read.csv(file='STAT101_1.csv')
dbWriteTable(conn = db, name = 'STAT101_1', 
             value = STAT101_1, row.names = FALSE)
rm(STAT101_1)#removes object from R's memory as it is now saved in a data base
STAT101_2 = read.csv(file='STAT101_2.csv')
dbWriteTable(conn = db, name = 'STAT101_2', 
             value = STAT101_2, row.names = FALSE)
rm(STAT101_2)
```
Let's inspect the results

```{r}
dbListFields(db, 'STAT101_1')
dbReadTable(db, 'STAT101_1')
dbReadTable(db, 'STAT101_2')
```
Now, if we want to run an analysis, we need to convert this relational data base back
into a flat file:
```{r}
STAT101_1 = dbReadTable(db, 'STAT101_1')
STAT101_2 = dbReadTable(db, 'STAT101_2')
(STAT101 = rbind(STAT101_1,STAT101_2))
```
Lastly, it is good practice to close connections after you are done with them:
```{r}
dbDisconnect(db)
```

